
---
output: md_document
title: "Introduction to PoDDyHePo package"

---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Introduction

The purpose of this package is to project prevalences of health indicators in future. The method uses data on cross-sectional surveys and national population forecasts to take into account the changing population structure. This package contains the following functions:

1. `PoDDyHePoFreqTable`: creates a descriptive table with sample size and amount of missingness in the data.
2. `PoDDyHePoModelSelection`: selects models for according to BIC for projection step.
3. `PoDDyHePoPopulationDF`: uses population forecasts from Statistics Finland and well-formatted survety data to create a data frame ready for projection.
4. `PoDDyHePoProjection`: projects individual-level values of the variable using multiple imputation. 
5. `PoDDyHePoPool`: caculates prevalences by pooling results from `PoDDyHePoProjection`.
6. `PoDDyHePoPlot`: plots the prevalences against survey year. 

# Getting started

Before getting started, make sure the following package has been attached: `splines`.

```{r attach_package, warning=FALSE, error=F, eval=F}
library(splines)
```

For all the examples, please load `testdata` (synthetic data built in the package):

```{r load_testdata, warning=FALSE, error=F, eval=F}
library(csprojections)
data(testdata)
```

In the `testdata`, we have mandatory variables: 

 * `year`: Survey year. Numeric. 1997, 2002, 2007, 2012 and 2017.
 * `sex`: Categorical. 1 for men and 2 for women.
 * `age`: Categorical. Ranges from 25 to 64.

Main variables of interest: 

 * `obe`: Obesity/Overweight/Normal. Categorical. 0 = Normal weight, 1 = Overweight and 2 = Obesity.
 * `smo`: Smoking status. Binary. 0 = Non-smoker, 1 = Smoker.
 * `highbp`: Hypertension. Binary. 0 = Non-hypertensive, 1 = hypertensive.
 * `dbt`: Diabetes. Binary. 0 = Non-diabetic, 1 = diabetic.
 * `sed`: Sedentary lifestyle. Binary. 0 = Non-sedentary lifestyle, 1 = sedentary lifestyle.
 * `highkol`: Elevated total cholesterol. Binary. 0 = Not high total cholesterol, 1 = high total cholesterol.

Other background variables: 

 * `edulv`: Education level. Categorical. 1 = Low education level, 2 = Middle education level, 3 = high education level.
 * `marsta`: Marital Status. Binary. 0 = Unmarried/Separated or Divorced/Widowed, 1 = Married/Cohabiting/Registered relationship.
 * `area`: Categorical. 2 = North Karelia, 3 = North Savonia, 4 = Turku and Loimaa, 5 = Helsinki and Vantaa, 6 = Northern Ostrobothnia and Kainuu.

## Guidelines for Using the Package

**Step 1**: If data from past surveys include missing values, they are imputed before selecting models for projection.

```{r create_multiple_imputed_data_set, warning=F, error=F, message=F, eval=F}
imp <- mice(testdata, m = 5, maxit = 10, printFlag = F)
```

NB: `m` and `maxit` can be any number you want, but please consider the computational power of your workstation. Usually, `m = 10` or `m = 20` should be enough. And if `printFlag = T`, `mice` will print history on console. Here, `printFlag=F` is used for silent computation. Data set should be well-formatted like `testdata`, which means there are mandatory variables: year, sex and age named exactly in this way, the main variables of interest whose prevalences to be estimated, and possibly other background variables. Data from different years are merged into the same data frame.

**Step 2**: Select models with PoDDyHePoModelSelection. The user can specify if some variables are allowed to have non-linear effects, which are modelled using natural splines. Some of the variables can also be forced (fixed) to be included in the model. Here in this example, we define knots and boundary knots in the spline function `ns`, and fix variables year, sex, age and an interaction term: year:sex.

In the function `PoDDyHePoModelSelection(imp, DV, NsVar = NULL, df = NULL, knots = NULL, b.knots = NULL, f.var = NULL)`, the first argument `imp` is an object of class mids from `mice()` in mice package (see *Step 1*). `DV` is the abbreviation of dependent variable. Arguments `NsVar`, `df`, `knots`, `b.knots` are for customizing `ns` function in the model. `NsVar` specifies a variable with a nature spline, if it is NULL, degrees of freedom(`df`), breakpoints that define the spline (`knots`) and boundary points (`b.knots`) are ignored; if it is not NULL, we can set df, knots and b.knots. Examples are given as follows based on `testdata`. More details about `ns` function can be found via https://www.rdocumentation.org/packages/splines/versions/3.6.2/topics/ns.

```{r mode_selection_1, warning=F, error=F, message=F, eval=F}
smo <- PoDDyHePoModelSelection(imp, 
                               NsVar = "year", 
                               DV = "smo", 
                               knots = 2002, 
                               b.knots = c(1997, 2017), 
                               f.var = c("year", "sex", "age", "year:sex"))
```

For other variables, if other arguments remain the same, what we need to do is to change `DV`, like

```{r mode_selection_2, warning=F, error=F, message=F, eval=F}
obe <- PoDDyHePoModelSelection(imp, 
                               NsVar = "year", 
                               DV = "obe", 
                               knots = 2002, 
                               b.knots = c(1997, 2017), 
                               f.var = c("year", "sex", "age", "year:sex"))
```

This function returns a list with 4 elements. One for recording the selection process, one for storing the mean BIC from each imputed data set, one for saving the selected models, and the last one is for the suggested model. Please notice that, the projections may be sensitive to the imputation models used, so it is recommended to carry out sensitivity analyses by trying alternative models. 

**Step 3**: Use population forecasts from Statistics Finland and our survey data (here is `testdata`) to create a data frame for projection.

```{r dataset_creation, warning=F, error=F, message=F, eval=F}
testdata2040 <- PoDDyHePoPopulationDF(testdata,
                                      file = "~path/003_139f_2040_20211102-100631.csv", 
                                      size = 10000, 
                                      y2pred = c(2021, 2025, 2030, 2035, 2040))
```

For the function `PoDDyHePoPopulationDF`, the first argument is the well-formatted data, in the example `testdata`. The second argument, which is "file", is a csv file downloaded from Statistics Finland (with special requirements given below). The third is a sample size and the fourth is the years to be projected. Above, 2021, 2025, 2030, 2035 and 2040, are the years to be projected, so `y2pred = c(2021, 2025, 2030, 2035, 2040)`.


In order to make this function work, the population forecast data has to be in a specific form. Data can be downloaded in the correct form by following the next steps: 
  
1. Visit https://pxnet2.stat.fi/PXWeb/pxweb/en/StatFin/StatFin__vrm__vaenn/, scroll down and find Population projection.
2. Click 139f -- Population projection 2021: Population according to age and sex by area, 2021-2040 [Size: 5818 Kb] [Modified: 9/30/2021].
3. Select Population 31 Dec (projection 2019) (Information) - WHOLE COUNTRY (Area) - Select the years to be projected (Year) -Select all (Sex) - Select all (Age), and click Show table. (In our example, it is 2021, 2025, 2030, 2035 and 2040)
4. Click "pivot manual" and set Area, Age in the Rows box (Order: Area, Age) and set Year, Sex, Information in the Columns box (Order: Year, Sex, Information) and click "complete".
5. Click "Save the result as ..." - select "Semicolon delimited with heading" and save.


**Step 4**: Project values for the future years.

```{r projection, warning=F, error=F, message=F, eval=F}
projection_with_knots <- PoDDyHePoProjection(testdata2040, 
                                             m = 5, 
                                             sep_col = "obe",
                                             smo ~ ns(year, knots = c(2002), Boundary.knots = c(1997, 2017)) + 
                                               sex + age + edulv + sed + marsta + obe + ns(year, knots = c(2002), Boundary.knots = c(1997, 2017)):sex + age:edulv + age:marsta +
                                               age:obe + sex:age
                                             obe$`Suggested Model`,
                                             dbt$`Suggested Model`,
                                             sed$`Suggested Model`,
                                             highbp$`Suggested Model`,
                                             highkol$`Suggested Model`,
                                             edulv$`Suggested Model`,
                                             marsta$`Suggested Model`,
                                             area$`Suggested Model`)

```

In the `PoDDyHePoProjection` function, the first argument is the data set from *Step 3*. The second argument is the number of multiple imputations, by default this is 5, but much greater value is usually needed in practice. The third argument is a categorical variable with three or more levels, like in the testdata, `obe` has three levels: 0 = Normal Weight, 1 = Overweight, 2 = Obesity. By specifying `sep_col = "obe"`, the function calculates the proportions of each category. The last part is for the formulas of the imputation models. In *Step 2*, we found models for variables with missingness, and they are saved in the `Suggested Model` in the list given by `PoDDyHePoModelSelection`. So, it can be used in a way like `obe`: obe$\`Suggested Model\`, or specified our own model like for `smo` in the code above. Every variable with missingness has to be given a formula. When there is a variable with 3 or more levels like `obe` in the example, it has to be specified in `sep_col`, as we use Wilson Confidence Interval in the pooling step, which is for binomial proportions. 

**Step 5**: Pool the results from the projection step and plot the prevalences.

```{r pool_and_plot, warning=F, error=F, message=F, eval=F}
# Smoking status
smopool <- PoDDyHePoPool(projection_with_knots, "smo")
PoDDyHePoPlot(smopool, year = 2017, "Smoking")

# Normal Weight/Overweight/Obesity
obepool <- PoDDyHePoPool(projection_with_knots, "obe")
PoDDyHePoPlot(obepool, year = 2017, "Normal Weight/Overweight/Obesity", sepvarlbl = c("Normal Weight", "Overweight", "Obesity"))

# Diabetes
dbtpool <- PoDDyHePoPool(projection_with_knots, "dbt")
PoDDyHePoPlot(dbtpool, year = 2017, "Diabetes")

# Sedentary Lifestyle
sedpool <- PoDDyHePoPool(projection_with_knots, "sed")
PoDDyHePoPlot(sedpool, year = 2017, "Sedentary")

# Hypertension
highbppool <- PoDDyHePoPool(projection_with_knots, "highbp")
PoDDyHePoPlot(highbppool, year = 2017, "Hypertension")

# Elevated Total Cholesterol
highkolpool <- PoDDyHePoPool(projection_with_knots, "highkol")
PoDDyHePoPlot(highkolpool, year = 2017, "Elevated Total Cholesterol")

```

In the function `PoDDyHePoPool`, the first argument is the data with projected values from *Step 4*, and the second argument is the main variable whose prevalences we aim to estimate. In the function `PoDDyHePoPlot`, the first argument is the result from `PoDDyHePoPool`; the second argument is the maximum year in the observed data (In `testdata`, it is 2017), the third argument is the title for the figure. When there is a variable has three or more levels, an extra argument `sepvarlbl` is available, which is to label the levels of this variable. Otherwise, the labels for the newly created variables will be `obe_0`, `obe_1` and `obe_2`.


## Descriptive Table
Descriptive table with sample size and amount of missingness is also available by using `PoDDyHePoFreqtable`.

There are two arguments in this function, the first one is the observed data (`testdata`) while the second argument is a variable name of interest. Examples are given below.

```{r descriptive_table, warning=F, error=F, message=F, eval=F}
PoDDyHePoFreqTable(testdata, colName = "smo")
PoDDyHePoFreqTable(testdata, colName = "obe")
```




*.Rproj
.Rproj.user
.Rhistory
.RData
.Ruserdata


^.*\.Rproj$
^\.Rproj\.user$
^.*\.log$
^README.Rmd
^csprojections_0.1.0.tar.gz

